---
title: "ADA Final Data Analysis"
author: "Patrick Cannon"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(ggplot2, tidyverse, ROCR, haven, readr, dplyr, forcats, table1, odds.n.ends, blorr, lmtest, car)
```

## Import and Clean BRFSS dataset (2020)
```{r import}
brfss <- read_xpt("LLCP2020.XPT")
```

## Data Preparation and Cleaning
```{r dependent variable}
## Dependent variable is coded CIMEMLOS, with question:
## "During the past 12 months, have you experienced confusion or memory 
## loss that is happening more often or is getting worse?"

table(brfss$CIMEMLOS, useNA = "always") # determine initial count of each type
```

```{r recode DV}
## make a binary cognitive decline variable that categorizes individuals into 
## yes or no and excludes other responses
brfss$cognitiveDecline_binary <- ifelse(brfss$CIMEMLOS == 2, 0, 
                                        ifelse(brfss$CIMEMLOS == 1, 1, NA))
```

```{r factor DV}
## make the binary into a factor variable
brfss$cognitiveDecline_binary <- factor(brfss$cognitiveDecline_binary, 
    levels = c(0,1), labels = c("No confusion or memory loss", 
                                "Yes, confusion or memory loss"))

label(brfss$cognitiveDecline_binary) <- "Does the respondent have confusion or memory loss in the past 30 days?"
```

```{r check DV recode}
## check to make sure re-classification worked
table(brfss$cognitiveDecline_binary, brfss$CIMEMLOS)
table(brfss$cognitiveDecline_binary, useNA = "always")
```

```{r drop NAs}
## omit NAs from dataset
brfss <- brfss[which(!is.na(brfss$cognitiveDecline_binary)),]
```

```{r recode main predictor}
## main predictor is sleep, so recode sleep
brfss <- brfss[!(brfss$SLEPTIM1 == 77 | brfss$SLEPTIM1 == 99),] #remove unwanted numeric values
brfss <- brfss[which(!is.na(brfss$SLEPTIM1)),] #also remove NAs
label(brfss$SLEPTIM1) <- "Hours of sleep gotten on average in the past 30 days"
```

```{r code race category}
# create race category

brfss<-rename(brfss, "race"="_RACE")
brfss$race[brfss$race==9] <- NA
brfss <- brfss[which(!is.na(brfss$race)),]


brfss <- brfss %>%
  mutate(race_cat = factor(race, 1:8, c("White, non-Hispanic", "Black, non-Hispanic", "American Indian or Alaksa Native, non-Hispanic", "Asian, non-Hispanic", "Native Hawaiian or other Pacific Islander, non-Hispanic", "Other race, non-Hispanic", "Multiracial, non-Hispanic", "Hispanic"))) #make race as a factor variable

label(brfss$race_cat) <- "Race/Ethnicity"
```

```{r recode diabetes}
## clean the other moderating variables
## diabetes
brfss$diabetes_binary<-ifelse(brfss$DIABETE4==3| brfss$DIABETE4==4, 0, ifelse(brfss$DIABETE4==1|brfss$DIABETE4==2, 1, NA))
brfss$diabetes_binary<-factor(brfss$diabetes_binary, levels=c(0,1), labels=c("No Diabetes", "Diabetes"))
brfss<-brfss[which(!is.na(brfss$diabetes_binary)),]
label(brfss$diabetes_binary) <- "Does the respondent have diabetes?"
```

```{r recode obesity}
## obesity
brfss$obesity_binary<-ifelse(brfss$"_RFBMI5"==1, 0, ifelse(brfss$"_RFBMI5"==2, 1, NA))
brfss$obesity_binary<-factor(brfss$obesity_binary, levels=c(0,1), labels=c("Not Overweight or Obese", "Overweight or Obese"))
brfss<-brfss[which(!is.na(brfss$"_RFBMI5")),]
label(brfss$obesity_binary) <- "Is the respondent obese or overweight?"
```

```{r recode smoking}
## smoking (at all)
brfss$smoking_binary<-ifelse(brfss$SMOKDAY2==3, 0, ifelse(brfss$SMOKDAY2==1 | brfss$SMOKDAY2==2, 1, NA))
brfss$smoking_binary<-factor(brfss$smoking_binary, levels=c(0,1), labels=c("Does not smoke", "Smokes"))
brfss<-brfss[which(!is.na(brfss$SMOKDAY2)),]
label(brfss$smoking_binary) <- "Does the respondent smoke at all"
```

```{r recode age}
## age
brfss<-rename(brfss, "age"="_AGEG5YR")
brfss$age[brfss$age==14] <- NA
brfss <- brfss[which(!is.na(brfss$age)),]

## recoding age to 13 categories as per BRFSS
brfss <- brfss %>%
  mutate(age_cat = factor(age, 0:12, c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80 or older"))) #make age as a factor variable based on cohort

label(brfss$age_cat) <- "Age cohort"
```

```{r recode sex}
## Sex
brfss$sex <- factor(brfss$SEXVAR, 
    levels = c(1,2), labels = c("Male", 
                                "Female"))
```

## Descriptives
## Table 1

```{r table 1}
table_1 <- table1(~ diabetes_binary + obesity_binary + smoking_binary + age_cat + race_cat + SLEPTIM1 + cognitiveDecline_binary | sex, data = brfss, topclass="Rtable1-zebra", render.missing=NULL)
table_1
```

## Analysis - Logistic Regressions

```{r initial LR with sleep and DV}
## Run a LR with number of hours slept as a numeric, continuous variable
sleep_continuous_logit <- glm(cognitiveDecline_binary ~ SLEPTIM1 , data = brfss, family = "binomial")
summary(sleep_continuous_logit)

odds.n.ends(sleep_continuous_logit, predProbPlot = TRUE) # look at summary values from model
```

```{r recode sleep into categories}
# Now, let's recode age to be in a range of 1-6, 7-9, and 10+ and see if there is a better prediction from there
# this is because our linearity assumption was not met
brfss$sleep_categories[(brfss$SLEPTIM1 >= 1 & brfss$SLEPTIM1 < 7)] <- 0
brfss$sleep_categories[(brfss$SLEPTIM1 >= 7 & brfss$SLEPTIM1 < 10)] <- 1
brfss$sleep_categories[(brfss$SLEPTIM1 >= 10 & brfss$SLEPTIM1 <= 24)] <- 2
```

```{r sleep categories}
## make new factor variable for sleep categories
brfss$sleep_categories <- factor(brfss$sleep_categories, levels = c(0:2), 
                                 labels = c("1 to 6 hours of sleep per night", 
                                            "7 to 9 hours of sleep per night", 
                                            "10 or more hours of sleep per night"))
```

```{r check sleep recode}
## check to see if recoding into factors worked
by(brfss$SLEPTIM1, brfss$sleep_categories, summary)
```

```{r re-run original LR}
## rerun the LR with sleep as categorical variable
sleep_categorical_logit <- glm(cognitiveDecline_binary ~ sleep_categories , data = brfss, family = "binomial")
summary(sleep_categorical_logit)

odds.n.ends(sleep_categorical_logit, predProbPlot = TRUE) # look at summary values from model
```

```{r add confounders}
## add age, sex, race, smoking, obesity, diabetes statuses as confounding variables
sleep_categorical_logit_adjusted <- glm(cognitiveDecline_binary ~ sleep_categories + age_cat + race_cat + smoking_binary + obesity_binary + diabetes_binary, data = brfss, family = "binomial")
summary(sleep_categorical_logit)

odds.n.ends(sleep_categorical_logit_adjusted, predProbPlot = TRUE) # look at summary values from model
```

```{r check for EM}
## incorporate an interaction term for sex and sleep categories to determine if there is any effect modification
sleep_categorical_logit_adjusted_em <- glm(cognitiveDecline_binary ~sleep_categories*SEXVAR + age_cat + race_cat + smoking_binary + obesity_binary + diabetes_binary, data = brfss, family = "binomial")
summary(sleep_categorical_logit)

odds.n.ends(sleep_categorical_logit_adjusted_em, predProbPlot = TRUE) # look at summary values from model
```

```{r male OR}
## look at male sex OR 
sleep_categorical_logit_adjusted_em_male <- glm(cognitiveDecline_binary ~sleep_categories + age_cat + race_cat + smoking_binary + obesity_binary + diabetes_binary, data = brfss[which(brfss$SEXVAR==1),], family = "binomial")
summary(sleep_categorical_logit)

odds.n.ends(sleep_categorical_logit_adjusted_em, predProbPlot = TRUE) # look at summary values from model
```

```{r female OR}
## look at OR for females
sleep_categorical_logit_adjusted_em_female <- glm(cognitiveDecline_binary ~sleep_categories + age_cat + race_cat + smoking_binary + obesity_binary + diabetes_binary, data = brfss[which(brfss$SEXVAR==2),], family = "binomial")
summary(sleep_categorical_logit)

odds.n.ends(sleep_categorical_logit_adjusted_em, predProbPlot = TRUE) # look at summary values from model
```

## Assumption testing

## All design assumptions were met. The outcome variable is a binary yes/no depdendent
## variable. Independence of observations was met through the study design itself of
## one response per person. Adequate sample size was met as the total number of samples was
## well over 1000 people.
## Data assumptions are tested below.

```{r linearity assumption}
# using Box-Tidwell to test linearity assumption for sleep
brfss <- brfss %>%
  mutate(sleep.times.logsleep = SLEPTIM1 * log(SLEPTIM1)) #create term to test linearity

boxTidwellSleep <- glm(cognitiveDecline_binary ~ sleep_categories + sleep.times.logsleep, data = brfss, family = "binomial") #Box Tidwell technique, test the assumption of linearity

summary(boxTidwellSleep)

## since the interaction term is significant, the assumption of linearity is violated.
```

```{r multicollinearity assumption}
## vif testing
vif(sleep_categorical_logit_adjusted_em)

## all VIF values are less than 4, which is acceptable to show that there is no MC.
```

```{r points of influence}
## influence
plot(sleep_categorical_logit_adjusted, which=4, id.n=5, col="red") 

## Cook's D cutoff=0.0015
cutoff <- 0.0015 

## identify observations with a Cook's Distance greater than cutoff
influential_points <- as.data.frame(cooks.distance(sleep_categorical_logit_adjusted)) %>%
  mutate(influential_points=row_number()) %>%
  filter(`cooks.distance(sleep_categorical_logit_adjusted)` > cutoff)


## exclude the values shown in the Cook's D plot, and see how the models compare.
sleep_categorical_logit_adjusted_influential <- update(sleep_categorical_logit_adjusted,subset=c(-influential_points$influential_points))
summary(sleep_categorical_logit_adjusted_influential)

## compare coefficients between models with and without influential observations
compareCoefs(sleep_categorical_logit_adjusted, sleep_categorical_logit_adjusted_influential) 
```

```{r goodness of fit tests}
#Various pseudo R squares, log likelihood, deviance, AIC, BIC
blr_model_fit_stats(sleep_categorical_logit_adjusted)

#Hosmer lemeshow goodness of fit test: a significant p value indicates a bad fit
blr_test_hosmer_lemeshow(sleep_categorical_logit_adjusted)
```

```{r LR test}
#likeilihood ratio test: compare two nested models
lrtest(sleep_categorical_logit_adjusted, sleep_categorical_logit_adjusted_em)
```




